---
title: "Reproducible Research - Course Project 1"
author: "Kevin H. Prescott"
date: "8/8/2022"
output: 
  html_document :
      theme: flatly
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading and preprocessing data
### Loading required packages
```{r packages, echo = TRUE, warning = FALSE, message = FALSE}
library(dplyr)
library(ggplot2)
library(lubridate)
```

### Loading data
We begin by loading the Activity dataset which will be used in this analysis

```{r load_data, echo = TRUE}
activity_data <- read.csv("C:\\Users\\opti1039\\OneDrive - The Toronto-Dominion Bank\\Desktop\\JohnHopkinsCourse\\Data\\activity.csv", header=TRUE)
```

### Data overview and preprocessing
The raw data contains the following variables and formats :
```{r show_head, echo = TRUE}
head(activity_data)
```

We have three variables :

* Steps
* Date
* Intervals

Based on the initial sample shown above, it seems like we have missing values. Looking at summaries by variables :


```{r show_summary_steps, echo = TRUE}
summary(activity_data$steps)
```
```{r show_summary_interval, echo = TRUE}
summary(activity_data$interval)
```
We will thus need to handle missing values in the *steps* variable and see how the *interval* one was assigned for missing values

One last thing to check before starting the analysis is the format of the *date* variable. We would want it to be in a date format so that we can easily use it in our analysis.

```{r date_format_check, echo = TRUE}
class(activity_data$date)
```
We convert the *date* variable to a date format

```{r date_format_change, echo = TRUE}
activity_data$date <- ymd(activity_data$date)
class(activity_data$date)
```

# Question 1 : What is the mean total number of steps taken per day ?

### Total daily steps taken
First, we compute the total number of steps per day and store it in a dataframe
```{r sum_daily_steps_df, echo = TRUE}
sum_day_steps <- activity_data %>% filter(!is.na(steps)) %>% group_by(date) %>% summarize(total_daily_steps = sum(steps, na.rm = TRUE))
sum_day_steps
```

### Plot total daily steps taken
```{r sum_daily_steps_plot, echo = TRUE}
sum_day_steps_plot <- ggplot(sum_day_steps, aes(total_daily_steps)) +
                      geom_histogram(fill = "blue", binwidth = 1000) +
                      labs(title = "Total Steps per Day", x = "Steps per day", y = "Frequency")
sum_day_steps_plot
```

### Mean and median of total number of steps taken per day
```{r stat_daily_steps, echo = TRUE}
mean_day_steps <- mean(sum_day_steps$total_daily_steps,na.rm=TRUE)
median_day_steps <- median(sum_day_steps$total_daily_steps,na.rm=TRUE)
out_stat <- as.data.frame(c(mean_day_steps,median_day_steps),row.names=c("Mean","Median"))
out_stat
```

# Question 2 : What is the average daily activity pattern ?

### Summarize data based on average steps grouped by intervals
```{r avg_int_steps_df, echo = TRUE}
avg_int_steps <- activity_data %>% filter(!is.na(steps)) %>% group_by(interval) %>% summarize(avg_int_steps = mean(steps))
avg_int_steps
```

### Plot time series based on intervals and average number of steps taken
```{r avg_int_steps_plot, echo = TRUE}
avg_int_steps_plot <- ggplot(avg_int_steps, aes(interval, avg_int_steps)) +
                      geom_line(color = "blue") +
                      labs(title = "Average Steps by Interval", x = "Interval", y = "Steps")
avg_int_steps_plot
```

### Find the interval that contains the maximum number of steps on average
```{r max_int_steps, echo = TRUE}
avg_int_steps %>% filter(avg_int_steps == max(avg_int_steps))
```


# Imputing missing values

In order to account for the missing values in the *steps* variable, we will replace the *N/As* by the average steps value by interval

```{r replace_NA_df, echo = TRUE}
activity_data_noNA <- activity_data
na <- is.na(activity_data_noNA$steps)
avg_int <- tapply(activity_data_noNA$steps, activity_data_noNA$interval, mean, na.rm=TRUE, simplify=TRUE)
activity_data_noNA$steps[na] <- avg_int[as.character(activity_data_noNA$interval[na])]
summary(activity_data_noNA$steps)
```
### Total daily steps taken without NAs
Now that missing values have been replaced by the average value by interval, we repeat the summarizing an plotting steps from the first question
```{r sum_daily_steps_noNA_df, echo = TRUE}
sum_day_steps_noNA <- activity_data_noNA %>% filter(!is.na(steps)) %>% group_by(date) %>% summarize(total_daily_steps = sum(steps, na.rm = TRUE))
sum_day_steps_noNA
```

### Plot total daily steps taken without NAs
```{r sum_daily_steps_noNA_plot, echo = TRUE}
sum_day_steps_noNA_plot <- ggplot(sum_day_steps_noNA, aes(total_daily_steps)) +
                      geom_histogram(fill = "blue", binwidth = 1000) +
                      labs(title = "Total Steps per Day (missing values replaced by average by interval)", x = "Steps per day", y = "Frequency")
sum_day_steps_noNA_plot
```
### Mean and median of total number of steps taken per day
```{r stat_daily_steps_noNA, echo = TRUE}
mean_day_steps_noNA <- mean(sum_day_steps_noNA$total_daily_steps,na.rm=TRUE)
median_day_steps_noNA <- median(sum_day_steps_noNA$total_daily_steps,na.rm=TRUE)

out_stat_noNA <- as.data.frame(c(mean_day_steps_noNA,median_day_steps_noNA),row.names=c("Mean","Median"))
out_stat_diff <- cbind(out_stat, out_stat_noNA)
names(out_stat_diff) <- c("Stat_w_NA","Stat_wo_NA")
out_stat_diff$diff <- out_stat_diff$Stat_wo_NA - out_stat_diff$Stat_w_NA
out_stat_diff
```

# Question 3 : Are there differences in activity patterns between weekdays and weekends

Using the dataset with missing values replaced, we start by adding a weekday/weekend indicator variable

```{r wkend_ind, echo = TRUE}
activity_data_noNA <- activity_data_noNA %>% 
                      mutate(weektype = ifelse(weekdays(date)=="Saturday" | weekdays(date)=="Sunday", "weekend", "weekday"))
head(activity_data_noNA)
```

### Summarize data based on average steps grouped by intervals and weektype
```{r avg_int_type_steps_df, echo = TRUE}
avg_int_type_steps <- activity_data_noNA %>% group_by(interval, weektype) %>% summarize(avg_steps = mean(steps))
avg_int_type_steps
```

### Plotting results separately for weekdays and weekends
```{r avg_int_type_steps_plot, echo = TRUE}
wktype_int_steps_plot <-  ggplot(avg_int_type_steps, aes(interval, avg_steps, color = weektype)) +
                          geom_line() +
                          facet_wrap(~weektype, ncol = 1, nrow=2) + 
                          labs(title = "Average Steps Taken by Interval", x = "Interval", y = "Steps")
wktype_int_steps_plot
```